Dark Control:\\ A Unified Account of Default Mode Function\\ by Markov Decision Processes

### 2017-09-10 paper/paper.tex
some more refs to FEP and FEP-RL links

\subsection{The free-energy principle and active inference}
% INTRO
According to theories of the free-energy principle (FEP) and
active inference \citep{friston2010free, fristonAIorRL} (see also \citep{dayan1995helmholtz}),
the brain corresponds to
a biomechanical reasoning engine. It is
dedicated to minimizing the long-term average of surprise: the log-likelihood of the observed sensory input --more precisely, an upper bound thereof-- \textit{relative} to the expectations about the external world derived from internal representations. The brain
would continuously generate hypothetical explanations of the world
and predict its sensory input.
% Precursors of this interpretation can be traced back to Dayan and colleagues \citep{dayan1995helmholtz}
% who introduced the \textit{Helmholtz machine}, a hierarchical factorial directional deep belief network. According to the FEP account,
% the goal of the brain is to optimize over a generative model $G$ of sensations: to iteratively
% modify its internal representation $p_G(\z|\x)$ of objects in the world, their interactions and dynamics. The model then allows
% to minimize surprise when these representations are confronted with sensory input $\x$ during action-perception cycles (i.e., the \textit{generative} model).
% The FEP also proposes a companion model called the \textit{recognition} model $p_R(\z|\x)$,
% which works in tandem with the generative model to accomplish approximate inference. Put differently, the recognition model dreams imaginary worlds $\z$,
% while the generative model tries to produce sensory sensations $\x \sim p_G(\x|\z)$ which match these
% imagined mental scenarios.
However, surprise is challenging to optimize numerically
because we need to sum over all hidden causes $\z$ of the sensations (an intractable problem).
Instead, FEP therefore minimizes an upper-bound on surprise given by

\begin{equation}
  \begin{split}
    \text{generative surprise } &:= -\log(p_G(\x)) = F_G(\x) \\
    &=\underbrace{F^R_G(\x)}_{\text{accuracy}} - \underbrace{\operatorname{{KL}}(p_R(\z|\x) || p_G(\z|\x))}_{\text{complexity}} \\
    &\le F^R_G(\x),
    \text{ with equality if }p_R(\z|\x) = p_G(\z|\x)\text{ for all } \z.
  \end{split}
\end{equation}
where
\begin{equation}
  F_G^R(\x) := \langle -\log(p_G(\z, \x))\rangle_{p_R(\z|\x)} - \mathcal H(p_R(\z|\x))
\end{equation}
is the \textit{free energy}.
Here, $\langle . \rangle$ is the \textit{expectation} operator, $\mathcal H$ is the \textit{entropy} functional while $\operatorname{KL}(.|.)$ is the usual \textit{Kullback-Leibler (KL) divergence},
which is a measure of how different two probability distributions are. In this framework, the goal of the agent is then to iterative refine the generative model $p_G$ and the recognition model $p_R$ so as to minimize the free energy $F_G^R(\x)$ over sensory input $\x$.
% The optimization procedure proposed by the authors derives from the \textit{wake-sleep algorithm} \citep{dayan1995helmholtz}.
% \begin{figure}[!tb]
%   \centering
%   \includegraphics[width=.8\linewidth]{vae.pdf}
%   \caption{Variational autoencoders...}
%   \label{fig:vae}
% \end{figure}

@@

\begin{eqnarray*}
  -Q^{\text{FEP}} \sim \underbrace{\log (p_G)}_{\text{FEP generative surprise}}.
    \end{eqnarray*}
% \paragraph{Comparison to our proposed theory.}
% On the surface, a common point between the FEP and our proposed RL-based framework
%   is placing the minimization of a surprise signal at the core of brain function.
%   Indeed in RL, surprise minimization is subsumed  by accurate prediction of
%   rewarding outcomes in the future. A ``free-energy'' agent is barely a biomechanical machine which has the tendency to resist undesired phase-transitions
%   \citep{friston2010free,fristonAIorRL,ortega2013thermodynamics}.
Such a surprise-guided reinforcement learning scheme has previously been advocated
under the equivalent framework of information compression \citep{schmidhuber2010,mohamed2015}.
Nevertheless, minimization of surprise quantities alone
may be insufficient to explain the diversity of behaviors observed in humans and other intelligent animals.
% The dark-room experiment, for example, provides a formidable argument against in favor of the insufficiency of the FEP and active inference theories to explain intelligent behavior in humans: when a human is put in a dark room, they don't just sit down in a corner forever, what would be the FEP-optimal thing to do; they actively search for the light switch, and start exploring the depths of the room, possibly looking for food or the way out!

>>>>

\subsection{The free-energy principle and active inference}
% INTRO
According to theories of the free-energy principle (FEP) and
active inference \citep{friston2010free, fristonAIorRL} (see also \citep{dayan1995helmholtz}),
the brain corresponds to
a biomechanical reasoning engine. It is
dedicated to minimizing the long-term average of surprise: the log-likelihood of the observed sensory input --more precisely, an upper bound thereof-- \textit{relative} to the expectations about the external world derived from internal representations. The brain
would continuously generate hypothetical explanations of the world
and predict its sensory input $\bf{x}$ (analogous to the state-action $(s, a)$ pair in an MDP
framework).
% Precursors of this interpretation can be traced back to Dayan and colleagues \citep{dayan1995helmholtz}
% who introduced the \textit{Helmholtz machine}, a hierarchical factorial directional deep belief network. According to the FEP account,
% the goal of the brain is to optimize over a generative model $G$ of sensations: to iteratively
% modify its internal representation $p_G(\z|\x)$ of objects in the world, their interactions and dynamics. The model then allows
% to minimize surprise when these representations are confronted with sensory input $\x$ during action-perception cycles (i.e., the \textit{generative} model).
% The FEP also proposes a companion model called the \textit{recognition} model $p_R(\z|\x)$,
% which works in tandem with the generative model to accomplish approximate inference. Put differently, the recognition model dreams imaginary worlds $\z$,
% while the generative model tries to produce sensory sensations $\x \sim p_G(\x|\z)$ which match these
% imagined mental scenarios.
However, surprise is challenging to optimize numerically
because we need to sum over all hidden causes $\z$ of the sensations (an intractable problem).
Instead, FEP therefore minimizes an upper-bound on surprise given by

\begin{equation}
  \begin{split}
    \text{generative surprise } &:= -\log(p_G(\x)) = F_G(\x) \\
    &=\underbrace{F^R_G(\x)}_{\text{accuracy}} - \underbrace{\operatorname{{KL}}(p_R(\z|\x) \| p_G(\z|\x))}_{\text{complexity}} \\
    &\le F^R_G(\x),
    \text{ with equality if }p_R(\z|\x) = p_G(\z|\x)\text{ for all } \z.
  \end{split}
\end{equation}
where
\begin{equation}
  F_G^R(\x) := \langle -\log(p_G(\z, \x))\rangle_{p_R(\z|\x)} - \mathcal H(p_R(\z|\x))
\end{equation}
is the \textit{free energy}.
Here, the angular brackets denote the expectation of the joint negative log-likelihood $-\log(p_G(\z, \x))$ w.r.t the recognition density $p_R(\z|\x)$, $\mathcal H$ is the \textit{entropy} functional defined by $\mathcal H(p) := -\sum_{\z}p(\z)\log(p(\z)$, while $\operatorname{KL}(.\|.)$ is the usual \textit{Kullback-Leibler (KL) divergence} defined by $\operatorname{KL}(p\|q) := \sum_{z}p(\z)\log(p(\z)/q(\z))$,
which is a measure of how different two probability distributions are. In this framework, the goal of the agent is then to iteratively refine the generative model $p_G$ and the recognition model $p_R$ (i.e two Bayesian belief nets) so as to minimize the free energy $F_G^R(\x)$ over sensory input $\x$. $F^R_G(x)$ is low just in case
\begin{itemize}
\item  $p_R(\z|\x)$ puts a lot of mass on configurations $(\z,\x)$ which are $p_G$-likely, % (i.e the sensation $\x$ is likely caused by $\z$)
  and
\item $p_R(\z|\x)$ as uniform as possible (i.e have high entropy), so as not to concentrate its mass only on a small subset of possible causes for the sensation $\x$.
  \end{itemize}
% The optimization procedure proposed by the authors derives from the \textit{wake-sleep algorithm} \citep{dayan1995helmholtz}.
% \begin{figure}[!tb]
%   \centering
%   \includegraphics[width=.8\linewidth]{vae.pdf}
%   \caption{Variational autoencoders...}
%   \label{fig:vae}
% \end{figure}

@@

\begin{eqnarray*}
  -Q^{\text{FEP}} \approx \underbrace{F^R_G(\x)}_{\text{negative free energy}} \approx \underbrace{-\log (p_G)}_{\text{FEP generative surprise}}.
    \end{eqnarray*}
% \paragraph{Comparison to our proposed theory.}
% On the surface, a common point between the FEP and our proposed RL-based framework
%   is placing the minimization of a surprise signal at the core of brain function.
%   Indeed in RL, surprise minimization is subsumed  by accurate prediction of
%   rewarding outcomes in the future. A ``free-energy'' agent is barely a biomechanical machine which has the tendency to resist undesired phase-transitions
%   \citep{friston2010free,fristonAIorRL,ortega2013thermodynamics}.
Such a surprise-guided reinforcement learning scheme has previously been advocated
under the equivalent framework of energy-based reinforcement-learning ~\citep{sallans2004,elfwing2016free} and information compression \citep{schmidhuber2010,mohamed2015}.
Nevertheless, minimization of surprise quantities alone
may be insufficient to explain the diversity of behaviors observed in humans and other intelligent animals.
% The dark-room experiment, for example, provides a formidable argument against in favor of the insufficiency of the FEP and active inference theories to explain intelligent behavior in humans: when a human is put in a dark room, they don't just sit down in a corner forever, what would be the FEP-optimal thing to do; they actively search for the light switch, and start exploring the depths of the room, possibly looking for food or the way out!
